The impact of gas bulk rotation on the Lyman-$\alpha$ line

### 2014-02-19 paper/ms.tex
My final version before submitted finally !!

\altaffiltext{1}{Departamento de F\'{i}sica, Universidad de los Andes, Cra. 1
No. 18A-10, Edificio Ip, Bogot\'a, Colombia}
\altaffiltext{2}{Max Planck Institute for Astrophysics, Karl-Schwarzschild-Str. 1, 85741, Garching, Germany}

@@

A basic understanding of the quantitative behaviour of the \ly line
has been reached through analytic studies in the case of a static
configurations, such as uniform slabs
\citep{Adams72,Harrington73,Neufeld90} and uniform spheres
\citep{Dijkstra06}. Analytic studies of configurations including
some kind of bulk flow only include the case of a sphere with a Hubble
like expansion flow \citep{LoebRybicki}.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.45\textwidth]{f5_1.eps}
  \includegraphics[width=0.45\textwidth]{f5_2.eps}
  \includegraphics[width=0.45\textwidth]{f5_3.eps}
  \includegraphics[width=0.45\textwidth]{f5_4.eps}    
\end{center}
    \caption{2D histogram of $N_{\rm scatt}$ vs $x$. The upper (lower) panels
      show the homogeneous (central) source distribution. Left
      corresponds to the static case and right
      $V_{\rm max}=300\kms$. The color scale is logarithmic on the
      number of photons with given values of $N_{\rm scatt}$ and
      $x$. \label{fig:Nscatt2D}}   
\end{figure*}

@@

In Figure \ref{fig:Nscatt} we show the average number of scatterings
$\langle N_{\rm scatt}\rangle$ as a function of the rotational velocity
$V_{\rm max}$. For the central source distribution the average number of
scatterings $\langle N_{\rm   scatt}\rangle$ changes less than $0.5\%$
for different velocities. In this set-up the number of scatterings is
proportional to the optical depth, with $\langle N_{\rm
  scatt}\rangle= (1.50, 1.00, 0.92)\tau_{\rm   H}$ for optical depth
values of $\tau_{\rm H} = (10^{5}, 10^{6}, 10^{7})$, respectively and
$V_{\mathrm{max}}=300$ \kms.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.90\textwidth]{f11.eps}
\end{center}
\caption{Triple peaked profiles for three different off-center positions
  $C=0.25R$ (doted line), $C=0.5R$ (dashed line) \& $C=0.75R$ (solid
  line) and for different rotational velocities. And $\theta$ represents the angle 
  between the rotation axes $\hat{k}$ and the observer line of sight.
   \label{fig:3p_profiles}} 
\end{figure*}

@@

Figure \ref{fig:OCspheres} presents the results for the $F(\mu)$
distribution for an sphere with $\tau_{\rm H}=10^{7}$ and $V_{\rm
  max}=200$\kms. The dashed and dotted lines represent the homogeneous
and central source distribution, respectively. This shows that the
the  anisotropy induced by rotation is at the $3\%$ level. We
performed the same test with all the models and find that the
variation in $\mu$ does not depend on the rotational velocity;
however, high optical depth values can increase the variation of
$\mu$ up to a $15\%$.

@@

Furthermore, taking into account all the models that do not event present hints for
a triple peaked emission it would seem that the rotational velocity
and optical depth must balance each other in order to show this feature
feature. Namely, if the rotational velocity is close to half the FWHM
of the line, the triple peaks are more likely to occur. This is the
same threshold that in the case of centered emission marks the
difference between double and single peaked lines.

@@

Another relevant sample of observations was presented by \cite{Yamada2012} 
for 91 LAEs at $z=3.1$. They find that $\sim 33 \%$ of the galaxies present 
a single peak profile. Some of the peaks have a strong symmetry, while
others do not. The symmetric single peaked profile can be explained by
rotation effects, while the asymmetric cases naturally arise in
outflow/inflow models \citep{Verhamme2008, Dijkstra06}. Among the 91
LAEs, 8 reveal triple peaked profiles with a central peak and two
smaller peaks at the redder and bluer side, a feature that is present
in our off-center emission models with rotation Figure \ref{fig:3p_profiles}.

The presence of single peaked profiles has been associated to
inflow/outflow dynamics \citep{Verhamme06,DijkstraKramer}. In this
paper we show that gas bulk rotation can also be considered as a
probable origin for that behaviour, provided that the observed single peak is highly
symmetric. Similarly, in the case of double peaked lines with a high
level of flux at the line center, rotation also deserves to be considered in
the pool of possible bulk flows responsible for that feature,
specially if the two peaks have similar intensities. The case of
triple peaked lines is another clear feature of gas rotation under the
additional condition of an off-centered emission.

@@

\item For homogeneous sources we find that the changes in the
line morphology are linked to the reduction in the average number of
scatterings. In the second part of \S \ref{sec:widthpeak} we show
that, as rotational velocity increases, a large fraction of photons
escape with less than $10$ scatterings. The average number of
scatterings decreases a $40\%$ compared to the static case. 
The reduction in the number of scattering decreases the probability
for a photon to move away from the line's center. This is also
translated into an increase of the escape fraction by a factor of
$\sim 2$ with respect to the static situation.

@@

>>>>

\altaffiltext{1}{Departamento de F\'{i}sica, Universidad de los Andes, Cra. 1
No. 18A-10, Edificio Ip, Bogot\'a, Colombia}
\altaffiltext{2}{Institute of Theoretical Astrophysics, University of Oslo, Postboks 1029, 0858 Oslo, Norway}

@@

A basic understanding of the quantitative behavior of the \ly line
has been reached through analytic studies in the case of a static
configurations, such as uniform slabs
\citep{Adams72,Harrington73,Neufeld90} and uniform spheres
\citep{Dijkstra06}. Analytic studies of configurations including
some kind of bulk flow only include the case of a sphere with a Hubble
like expansion flow \citep{LoebRybicki}.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.45\textwidth]{f5_1.eps}
  \includegraphics[width=0.45\textwidth]{f5_2.eps}
  \includegraphics[width=0.45\textwidth]{f5_3.eps}
  \includegraphics[width=0.45\textwidth]{f5_4.eps}    
\end{center}
    \caption{2D histogram of $N_{\rm scatt}$ vs $x$. The upper (lower) panels
      show the homogeneous (central) source distribution. Left panels 
      corresponds to the static case and right panels corresponds to
      $V_{\rm max}=300\kms$. The color scale is logarithmic on the
      number of photons with given values of $N_{\rm scatt}$ and
      $x$. \label{fig:Nscatt2D}}   
\end{figure*}

@@

In Figure \ref{fig:Nscatt} we show the average number of scatterings
$\langle N_{\rm scatt}\rangle$ as a function of the rotational velocity
$V_{\rm max}$. For the central source distribution the average number of
scatterings $\langle N_{\rm   scatt}\rangle$ changes less than $0.5\%$
for different velocities. In this set-up the number of scatterings is
proportional to the optical depth, with $\langle N_{\rm
  scatt}\rangle= (1.50, 1.00, 0.92)\tau_{\rm   H}$ for optical depth
values of $\tau_{\rm H} = (10^{5}, 10^{6}, 10^{7})$ respectively and
$V_{\mathrm{max}}=300$ \kms.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.90\textwidth]{f11.eps}
\end{center}
\caption{Triple peaked profiles for three different off-center positions
  $C=0.25R$ (doted line), $C=0.5R$ (dashed line) \& $C=0.75R$ (solid
  line) and for different rotational velocities. And $\theta$ represents the usual
  polar angle.
   \label{fig:3p_profiles}} 
\end{figure*}

@@

Figure \ref{fig:OCspheres} presents the results for the $F(\mu)$
distribution for a sphere with $\tau_{\rm H}=10^{7}$ and $V_{\rm
  max}=200$\kms. The dashed and dotted lines represent the homogeneous
and central source distribution, respectively. This shows that the
 anisotropy induced by rotation is at the $3\%$ level. We
performed the same test with all the models and find that the
variation in $\mu$ does not depend on the rotational velocity;
however, high optical depth values can increase the variation of
$\mu$ up to a $15\%$.

@@

Furthermore, taking into account all the models that do not present hints for
a triple peaked emission it would seem that the rotational velocity
and optical depth must balance each other in order to show this feature.
 Namely, if the rotational velocity is close to half the FWHM
of the line, the triple peaks are more likely to occur. This is the
same threshold that in the case of centered emission marks the
difference between double and single peaked lines.

@@

Another relevant sample of observations was presented by \cite{Yamada2012} 
for 91 LAEs at $z=3.1$. They find that $\sim 33 \%$ of the galaxies present 
a single peak profile. Some of the peaks have a strong symmetry, while
others do not. The symmetric single peaked profile can be explained by
rotation effects, while the asymmetric cases naturally arise in
outflow/inflow models \citep{Ahn2014,Verhamme2008, Dijkstra06}. Among the 91
LAEs, 8 reveal triple peaked profiles with a central peak and two
smaller peaks at the redder and bluer side, a feature that is present
in our off-center emission models with rotation Figure \ref{fig:3p_profiles}.

The presence of single peaked profiles has been associated to
inflow/outflow dynamics \citep{Verhamme06,DijkstraKramer}. In this
paper we show that gas bulk rotation can also be considered as a
probable origin for that behavior, provided that the observed single peak is highly
symmetric. Similarly, in the case of double peaked lines with a high
level of flux at the line center, rotation also deserves to be considered in
the pool of possible bulk flows responsible for that feature,
specially if the two peaks have similar intensities. The case of
triple peaked lines is another clear feature of gas rotation under the
additional condition of an off-centered emission.

@@

\item For homogeneous sources we find that the changes in the
line morphology are linked to the reduction in the average number of
scatterings. In the second part of \S \ref{sec:widthpeak} we show
that, as rotational velocity increases, a large fraction of photons
escape with less than $10$ scatterings. The average number of
scatterings decreases a $40\%$ compared to the static case. 
The reduction in the number of scatterings decreases the probability
for a photon to move away from the line's center. This is also
translated into an increase of the escape fraction by a factor of
$\sim 2$ with respect to the static situation.

@@

The data and source code and instructions to
replicate the results of this paper can be found
here {\texttt{https://github.com/jngaravitoc/RotationLyAlpha}}.
Thanks to the IPython community \citep{IPython} and to the matplotlib
community \citep{matplotlib}.
